En garde! Unguarded iteration for reversible computation in the Delay monad\thanks{Niccol\`o Veltri was supported by a research grant (13156) from VILLUM FONDEN.}

### 2019-04-30 mpc19/paper.tex

\section{Conclusions}

It is natural to wonder if our work can be ported to other monads of
partiality in Martin-L\"of type theory. As already discussed in the
end of Section \ref{sec:trace}, the maybe monad is not suitable for
modelling a well-behaved trace combinator without the assumption of
classical principles such as \textsf{LPO}. The partial map classifier
\cite{UustaluV17-2,EscardoK17}
$\mathsf{PMC}\,A = (P : \mathsf{Prop}) \to P \to A$, where
$\mathsf{Prop}$ is the type of propositions (types with at most one
inhabitant), supports the existence of a uniform iteration operator
and therefore a trace. Nevertheless, the specification of iteration is
more complicated then the one presented in Section \ref{sec:elgot} for
the delay monad, which is a simple corecursive definition. The Elgot
monad structure of $\mathsf{PMC}$ follows from its Kleisli category
being a join restriction category, so iteration is defined in terms of
least upper bounds of certain chains of morphisms. The subcategory of
partial isomorphisms of the Kleisli category of $\mathsf{PMC}$
supports a dagger trace combinator, which can be proved following
Kaarsgaard et al.'s general strategy \cite{KaarsgaardAG17}.  The exact
same observations apply to the partiality monad in homotopy type
theory \cite{AltenkirchDK17,ChapmanUV19}, to which the quotiented
delay monad $\mathsf{Delay}_{\approx}\,A = \Delay A/{\approx}$ is
isomorphic to under the assumption of countable choice.

>>>>

\section{Conclusions}
In this paper, we have extended the work of Carette and Sabry~\cite{CaretteS16}
to a (fully formalized) two-level calculus of \Pio{} programs and program
equivalences. Key in this effort was the use of the Kleisli category of the
$\Delay$ monad on $\Set$ under weak bisimilarity, which turned out to support
iteration via a trace that preserves all partial isomorphisms, in this way
giving semantics to the dagger trace of \Pio. Further, the work was formalized
using Agda 2.6.0.

It is natural to wonder if our work can be ported to other monads of partiality
in Martin-L\"of type theory. As already discussed in the end of Section
\ref{sec:trace}, the $\Maybe$ monad is not suitable for modelling a
well-behaved trace combinator without the assumption of classical principles
such as \textsf{LPO}. The partial map classifier \cite{UustaluV17-2,EscardoK17}
$\mathsf{PMC}\,A = (P : \mathsf{Prop}) \to P \to A$, where $\mathsf{Prop}$ is
the type of propositions (types with at most one inhabitant), supports the
existence of a uniform iteration operator and therefore a trace. Nevertheless,
the specification of iteration is more complicated then the one presented in
Section \ref{sec:elgot} for the delay monad, which is a simple corecursive
definition. The Elgot monad structure of $\mathsf{PMC}$ follows from its
Kleisli category being a join restriction category, so iteration is defined in
terms of least upper bounds of certain chains of morphisms. The subcategory of
partial isomorphisms of the Kleisli category of $\mathsf{PMC}$ supports a
dagger trace combinator, which can be proved following the general strategy in
\cite{KaarsgaardAG17}. The exact same observations apply to the partiality
monad in homotopy type theory \cite{AltenkirchDK17,ChapmanUV19}, to which the
quotiented delay monad $\mathsf{Delay}_{\approx}\,A = \Delay A/{\approx}$ is
isomorphic to under the assumption of countable choice.

Though the Kleisli category of the $\Delay$ monad on $\Set$ is well studied,
comparatively less is known about this monad on other categories. It could be
interesting to study under which conditions its iterator exist -- e.g., whether
this is still the case when $\Set$ is replaced with an arbitrary topos. Another
avenue concerns the study of \emph{time invertible} programming languages:
Though not immediately clear in the current presentation, the trace on
$\Inv\Dapprox$ is not just reversible (in the sense that it preserves partial
isomorphisms) but in fact time invertible, in the sense that the number of
computation steps needed to perform $\trace (\dagger f)$ is \emph{precisely}
the same as what is needed to perform $\trace~f$ on any input. Since the
$\Delay$ monad conveniently allows the counting of computation steps, we
conjecture that this is an ideal setting in which to study such intentional
semantics of reversible programming languages.
